---
title: "BST 270 In-class Project"
output: html_document
---

## Introduction
In this Rmarkdown file we will attempt to reproduce the figures, tables and analyses presented in the paper [_Relation between Optimism and Lipids in Midlife_.](https://www.ajconline.org/article/S0002-9149(13)00388-3/pdf)

1. Boehm, J. K., Williams, D. R., Rimm, E. B., Ryff, C., & Kubzansky, L. D. (2013). Relation between Optimism and Lipids in Midlife. The American Journal of Cardiology, 111(10), 1425-1431.
http://doi.org/10.1016/j.amjcard.2013.01.292

In 1995, MIDUS survey data were collected from a total of 7,108 participants. The baseline sample was comprised of individuals from four subsamples: (1) a national RDD (random digit dialing) sample ($n = 3,487$); (2) oversamples from five metropolitan areas in the U.S. ($n = 757$); (3) siblings of individuals from the RDD sample ($n = 950$); and (4) a national RDD sample of twin pairs ($n = 1,914$). All eligible participants were non-institutionalized, English-speaking adults in the contiguous United States, aged 25 to 74. All respondents were invited to participate in a phone interview of approximately 30 minutes in length and complete 2 self-administered questionnaires (SAQs), each of approximately 45 pages in length. In addition, the twin subsample was administered a short screener to assess zygosity and other twin-specific information. With funding provided by the National Institute on Aging, a longitudinal follow-up of MIDUS I began in 2004. Every attempt was made to contact all original respondents and invite them to participate in a second wave of data collection. Of the 7,108 participants in MIDUS I, 4,963 were successfully contacted to participate in another phone interview of about 30 minutes in length. MIDUS II also included two self-administered questionnaires (SAQs), each of about 55 pages in length, which were mailed to participants. The overall response rate for the SAQs was 81\%. Over 1,000 journal articles have been written using MIDUS I and II data since 1995.

Here we attempt to reproduce the findings of [1] and critique the reproducibility of the article. This particular article focuses only on MIDUS II data, including biomarker data, and investigates the relationship between optimism and lipids. The MIDUS II data and supporting codebook and other documents can be downloaded \href{https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4652?archive=ICPSR&q=MIDUS+}{here}. The data can be downloaded in multiple formats. Here we use the R data files and perform all data cleaning and analyses in R version 3.5.1. The biomarker data can be downloaded \href{https://www.icpsr.umich.edu/icpsrweb/NACDA/studies/29282}{here}.


## Data Dictionary
This manuscript uses several variables from multiple data files. Some of these variables don't have intuitive names and need to be manually looked up either online or in the codebooks provided in the data downloads. [This google sheet](https://docs.google.com/spreadsheets/d/1gMIepQtsMXc0g2cJChEmmeSR2ltlfwwUZ-Lo2HU_WPg/edit?usp=sharing) acts as a data dictionary for only the variables we believe were used for the analysis according to our understanding of the methods section of the paper.

## Load needed packages
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(arsenal)
library(kableExtra)
```


## Load and combine data
```{r}
load("ICPSR_04652/DS0001/04652-0001-Data.rda") # load MIDUS2 interview/questionnaire data
load("ICPSR_29282/DS0001/29282-0001-Data.rda") # load MIDUS2 biomarker data
df <- inner_join(da29282.0001, da04652.0001, by="M2ID") # merge the two dataframes using an inner join
dim(df)
```
We have a total of 1,054 participants after merging the data frames. 

### Optimism Score
We can drop any observation with a missing value in the `B1SORIEN` column which indicates the overall optimism score for a given participant.

```{r}
df <- df %>% drop_na(B1SORIEN) # drop rows with missing optimism values 
dim(df)
```
We lose 4 participants.

### Lipid Measurments Filter

Next, we drop the participants with missing lipid level measurements. The authors state that they are considering "a lipid panel of total cholesterol, HDL cholesterol, LDL cholesterol, and triglycerides". Therefore, we can drop rows with missing values in the `B4BCHOL` (total cholesterol), `B4BLDL` (LDL cholesterol), `B4BHDL` (HDL cholesterol), and `B4BTRIGL` (triglycerides) columns.

```{r}
df <- df %>% drop_na(B4BCHOL, B4BLDL, B4BHDL, B4BTRIGL) # drop rows with missing lipid measurements
dim(df)
```
We lose 11 participants. 

### Potential Confounders Filter

#### Blood pressure

There are several columns for blood pressure medication, including `B1PA24B`, `B1PA24C`, `B1PA24D`, and `B4XPC_1156`. Using any of these variables for our filter leads to less than 990 participants. This may be because those without high blood pressure did not answer a question about whether or not they used medication to control "their" high blood pressure, thus leading to a lot of missing values. This appears to be the case when consulting the data cookbooks as well, so we will have to recode NAs to 2 (NO) if they don't have high blood pressure (found in `B1PA24`). We choose to use `B1PA24B` as our blood pressure indicator as well, as it has the least missing values.


```{r}
df <- df %>% drop_na(B1PA24) # drop rows with missing diagnosis of blood pressure
dim(df)

# fix blood pressure column as described above; if they answered no to ever being diagnosed with high blood pressure
# and have an NA for ever taking blood pressure medication, change NA to 2 (no).
df[is.na(df$B1PA24B),]$B1PA24B <- ifelse(df[is.na(df$B1PA24B),]$B1PA24 == "(2) No", "(2) No", NA)

table(df$B1PA24B)
df <- df %>% drop_na(B1PA24B)
dim(df)
```

We lose 1 individual.

#### Chronic conditions

Similar to blood pressure above, we will use `B4H1A`, `B4H1B`, `B4H1F`, `B4H1I` to replace some of the NAs found in `B4H1AD`, `B4H1BD`, `B4H1FD`, `B4H1ID`, since the data does not contain 9 as stated in the online variable information. 

```{r}
# Replace some NAs with 0s
df[is.na(df$B4H1AD),]$B4H1AD <- ifelse(df[is.na(df$B4H1AD),]$B4H1A == "(2) NO", "(2) NO", NA)
df[is.na(df$B4H1BD),]$B4H1BD <- ifelse(df[is.na(df$B4H1BD),]$B4H1B == "(2) NO", "(2) NO", NA)
df[is.na(df$B4H1FD),]$B4H1FD <- ifelse(df[is.na(df$B4H1FD),]$B4H1F == "(2) NO", "(2) NO", NA)
df[is.na(df$B4H1ID),]$B4H1ID <- ifelse(df[is.na(df$B4H1ID),]$B4H1I == "(2) NO", "(2) NO", NA)

# Drop rows that still contain NAs
df <- df %>% drop_na(B4H1AD, B4H1BD, B4H1FD, B4H1ID)

dim(df)
```

We lose 12 individuals.



Figure
```{r}
library(ggplot2)
ggplot(df, aes(x=B1SORIEN))+geom_histogram(bins=30)
```

